Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
section 3

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

The dichotomy between evaluation (and resource usage) of a case expression
whose scrutinee is in weak head normal form and one whose scrutinee is not
leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
on weak head normal formed-ness to accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
\end{array}
\]
The first function uses $x$ linearly, but the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee and all its used resources; or
constructor arguments and the resources used to type them.
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

@@

First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement, introducing
the case binder $z$ in the unrestricted environment as a $\D$-bound
variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
linear typing environment of the alternative, similarly to how the resources
used to type a $Let$ binder are still available in the continuation.
%
%The $\vdash_{alt}$ judgement is 
%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
%to type the scrutinee sub-expressions,
%and the name of the case binder $z$.
%
%
Although the main idea for typing ``WHNF case expressions'' is conveyed by
this rule, the full rule of our system is slightly more involved, as will be seen after
discussing cases of scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
modes that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

@@

Lastly, consider an alternative matching on a data constructor without any
linear components. The linear resources used to type a scrutinee matching such
a pattern are fully consumed during evaluation: the resulting unrestricted
constructor application can only be well-typed under an empty linear
environment.
%
Consequently, the resources to type the scrutinee which are carried over to the
case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
(see the latter below), must be reactively destroyed from branches where the
pattern is unrestricted. In failing to do this, we would allow the fully-consumed
resources to be used again in the branch.
% Really, in the WHNF case it is not really necessary to remove the things
% since they must already be empty be definition.
%
%This definition agrees with the intuition we have developed by example in the
%previous section, and with the typing rule we devised for alternatives matching
%constructors without linear components.
%
%Taking into account that case expressions introduce the linear resources of the
%scrutinee in the typing environment of all alternatives, and in the usage
%environment of the case binder, we must reactively update the typing
%environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF status of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
the linear environment typing the scrutinee), this latter substitution makes
the case-binder unrestricted in branches with unrestricted patterns.
%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.

@@

\subsubsection{Proof irrelevant resources}\label{sec:irrev}

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$:
\begin{enumerate}

@@

    Considering (1), we must resort to a more uniform method of guaranteeing we
    ``finish consuming'' the scrutinee. There are only two ways of
    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:

        \begin{itemize}
            \item The case binder, which refers to the whole result of evaluating the scrutinee;
            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
        \end{itemize}

@@

To encode  linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not
applicable) we introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$.
Proof irrelevant resources
are linear resources in every sense, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded or used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, and annotating the judgement with the proof irrelevant resources
for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requires the resources in the case binder, typing environment, and
judgement annotation to be made ``irrelevant''.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed. We uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced proof irrelevant
resources, we can now present the full $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

@@

As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
can be cleanly divided amongst the various sub-expressions of a constructor
application and, henceforth, each pattern variable, there is no direct mapping
between the resources typing a scrutinee not in WHNF and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce that all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having uniquely tagged the resources in the usage environment of each pattern
variable, we need only to express that (1) the pattern-bound $\D$-variables can
be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if a $K$-pattern-bound $\D$-variable is consumed, all remaining $K$-$\D$-variables,
standing for the remaining linear components of the same pattern, must also be
consumed.

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.

@@

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context (and the unrestricted context updated accordingly).

>>>>

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely \ROUNDTWO{discriminating by} the WHNF-ness of
scrutinee, irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

The dichotomy between evaluation (and resource usage) of a case expression
whose scrutinee is in weak head normal form and one whose scrutinee is not
leads to one of the key insights of $\lambda^\pi_\Delta$:
\ROUNDTWO{accurate typing of case expressions depends on whether the scrutinee is in WHNF.}
%
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
\end{array}
\]
The first function uses $x$ linearly, but the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee and all its used resources; or
constructor arguments and the resources used to type them.
\ROUNDTWO{The reader may wonder about the significance of case expressions
  where the scrutinee is in WHNF. We note that in an
  optimizing compiler (such as GHC) many intermediate programs that
  arise due to program transformations commonly scrutinize expressions
  in WHNF.}

@@

First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
introducing
the case binder $z$ in the unrestricted environment as a $\D$-bound
variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
linear typing environment of the alternative, similarly to how the resources
used to type a $Let$ binder are still available in the continuation.
%
%The $\vdash_{alt}$ judgement is 
%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
%to type the scrutinee sub-expressions,
%and the name of the case binder $z$.
%
%
Although the main idea for typing ``WHNF case expressions'' is conveyed by
this rule, the full rule of our system is slightly more involved, as will be seen after
discussing cases of scrutinees not in WHNF.

The alternative judgement \ROUNDTWO{$\G;\D \vdash \rho \Rightarrow e :^z_\D 
\s$} is used to type case alternatives, encompassing three
modes that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

@@

Lastly, consider an alternative matching on a data constructor without any
linear components. The linear resources used to type a scrutinee matching such
a pattern are fully consumed during evaluation: the resulting unrestricted
constructor application can only be well-typed under an empty linear
environment.
%
Consequently, the resources to type the scrutinee which are carried over to the
case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
(see the latter below), must be reactively destroyed from branches where the
pattern is unrestricted. In failing to do this, we would allow the fully-consumed
resources to be used again in the branch.
% Really, in the WHNF case it is not really necessary to remove the things
% since they must already be empty be definition.
%
%This definition agrees with the intuition we have developed by example in the
%previous section, and with the typing rule we devised for alternatives matching
%constructors without linear components.
%
%Taking into account that case expressions introduce the linear resources of the
%scrutinee in the typing environment of all alternatives, and in the usage
%environment of the case binder, we must reactively update the typing
%environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF status of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
to delete $\D_s$ from the usage environment of the \ROUNDTWO{$\D$-bound variable} $z$ in $\G$.
Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
the linear environment typing the scrutinee), this latter substitution makes
the case-binder unrestricted in branches with unrestricted patterns.
%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.

@@

\subsubsection{Irrelevant resources}\label{sec:irrev}

When the scrutinee is not in WHNF
\ROUNDTWO{we cannot statically determine which resources are consumed
  upon scrutinee evaluation or which are transferred into the branch continuation}
e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$:
\begin{enumerate}

@@

@@

\ROUNDTWO{
    Considering (1), we must resort to a more uniform method of guaranteeing we
    ``finish consuming'' the scrutinee. There are only two ways of
    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:

        \begin{itemize}
            \item The case binder, which refers to the whole result of evaluating the scrutinee;
            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
        \end{itemize}}

To encode  linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not
applicable) we introduce \emph{irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and \ROUNDTWO{pointwise} lifted to contexts
\ROUNDTWO{and usage environments} as $[\D]$.
Irrelevant resources
are linear resources in every sense, meaning they must be used
exactly once. However, since irrelevant resources cannot be
discarded or used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and the case
alternatives by introducing the case binder with an \ROUNDTWO{irrelevant} usage environment $[\D]$,
having the same irrelevant $[\D]$ in the typing
environment, and annotating the judgement with the irrelevant resources
for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requires the resources in the case binder, typing environment, and
judgement annotation to be made ``irrelevant''.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash \rho \Rightarrow_{\textsf{WHNF}}
e :^z_{\D_s} \s$ judgement is only well-defined for patterns $\rho$
matching the WHNF form 
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed. We uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced irrelevant
resources, we can now present the full $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
\ROUNDTWO{where the matching pattern is typed as before, with the
  case-binder $z$ is annotated with the appropriate usage environment;
  and the non-matching patterns are if the scrutinee were notin WHNF.}
We again note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

@@

As opposed to scrutinees in WHNF, where resources 
can be cleanly divided amongst the various sub-expressions of a constructor
application and, henceforth, each pattern variable, there is no direct mapping
between the resources typing a scrutinee not in WHNF and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce that all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given \ROUNDTWO{irrelevant} resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-bound variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having uniquely tagged the resources in the usage environment of each pattern
variable, we need only express that (1) the pattern $\D$-bound variables can
be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if a $K$-pattern $\D$-bound variable is consumed, all remaining
$K$-$\D$-bound variables,
standing for the remaining linear components of the same pattern, must also be
consumed.

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see~\cite{techreport}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.

@@

\paragraph{$\Delta$-bound variables} % Our development relies on a characterization of $\D$-bound
% variables
A well-typed program with a linear variable ($\xl$) is equivalently
well-typed if the linear variable were instead $\D$-bound ($\xD$) with
usage environment $\D$, with $\D$ available in the linear context and
\ROUNDTWO{with any occurrences of $x$ in usage environments in $\G$
  expanded into $x$'s usage environment $\Delta$, with the abuse of notation $\Gamma[\Delta/x]$.
